Future challenges for Research Software Engineering

### 2025-05-19 future_challenges_for_RSEng.tex
Edit section 2.3 to improve pagination

\subsection{Impact of AI and automation}
\label{sec:ai}
\begin{whatis}{}{impact-ai-automation}
RSEs understand how AI models work, and can help researchers reap their benefits, while
clearly knowing their limits.
\end{whatis}
Generative AI tools, such as GitHub Copilot \cite{Friedmann2021}, are being added to software development tools.
These AI powered code assistants promise to increase productivity by automatically generating code, and writing test cases and documentation.
Software developers use them to reduce the number of keystrokes, complete tasks and to recall syntax without having to consult the documentation \cite{Liang2024}.
Code assistants have the potential to help novice or even non-expert programmers write code \cite{Feldman2024} thus enabling non-experts to participate in research and software creation, and fostering greater collaboration across disciplines.
However, when these tools are used in practice, many users report increased cognitive load and frustration stemming from the difficulty to prompt the system effectively and from debugging the generated code \cite{Simkute2025}.
The potential of AI to be a democratising technology can only be realised when the algorithmic divide along with the digital divide is addressed \cite{Yu2020}.

@@

The lessons learned from Human Factors research of the previous decades \cite{Simkute2025} applied to AI tools helped resolve the initial issues outlined above.
We envision that AI will be transforming research processes by enabling rapid analysis of massive datasets, uncovering patterns and insights that would have been impossible or prohibitively time-consuming for humans alone to find.
In research and development, AI will accelerate scientific breakthroughs in fields like healthcare, climate science, and materials engineering by automating complex modelling and predictive tasks.
Developers will be shifting from manual coding to orchestrating AI-driven workflows, focusing more on high-level design, integration, and creative problem-solving while AI handles repetitive or routine tasks.

As AI continues to evolve, it will further blur the lines between human and machine roles, making research and software development faster.
With the general availability of AI, the interaction of RSEs with researchers will change at a fundamental level.
Whereas previously RSEs have been points of knowledge that could translate human language formulated problems into technical descriptions and then provide solutions to them,
it can be foreseen that AI will drive this low-key interaction to a highly specialised level of knowledge.

@@

\subsection{Impact of Complexity and Usability in Research Software}
\begin{whatis}{}{abstraction}
With a palette of readily reusable software components,
and with their skills in managing hierarchical systems of complex software components with clear interfaces,
RSEs can now build reliable, maintainable systems of a scale and capability not previously possible.
\end{whatis}

>>>>

\subsection{Impact of AI and automation}
\label{sec:ai}
\begin{whatis}{}{impact-ai-automation}
RSEs understand how AI models work, and can help researchers reap their benefits, while
clearly knowing their limits.
\end{whatis}
Generative AI tools, such as GitHub Copilot \cite{Friedmann2021}, are being added to software development tools.
These AI-powered code assistants promise increased productivity by automatically generating code, tests, and documentation.
Software developers use them to reduce the number of keystrokes, draft tasks, and to recall syntax without having to consult the documentation \cite{Liang2024}.
Code assistants can help novice or even non-expert programmers write code \cite{Feldman2024}, thus enabling non-experts to participate in research and software creation, and fostering greater collaboration across disciplines.
However, when these tools are used in practice, many users report increased cognitive load and frustration stemming from the difficulty to prompt the system effectively and from debugging the generated code \cite{Simkute2025}.
The potential of AI to be a democratising technology can only be realised when the algorithmic divide along with the digital divide is addressed \cite{Yu2020}.

@@

The lessons learned from Human Factors research of the previous decades \cite{Simkute2025}, applied to AI tools, helped resolve the initial issues outlined above.
We envision that AI will be transforming research processes by enabling rapid analysis of massive datasets, uncovering patterns and insights that would have been impossible or prohibitively time-consuming for humans alone to find.
In research and development, AI will accelerate scientific breakthroughs in fields like healthcare, climate science, and materials engineering by automating complex modelling and predictive tasks.
Developers will be shifting from manual coding to orchestrating AI-driven workflows, focusing more on high-level design, integration, and creative problem-solving while AI handles repetitive or routine tasks.

As AI continues to evolve, it will further blur the lines between human and machine roles.
With the general availability of AI, the interaction of RSEs with researchers will fundamentally change.
Whereas previously RSEs have been points of knowledge that could translate human language formulated problems into technical descriptions and then provide solutions to them,
we foresee that AI will drive this low-key interaction to a highly specialised level of knowledge.

@@

\vspace{-4ex}
\subsection{Impact of Complexity and Usability in Research Software}
\begin{whatis}{}{abstraction}
With a palette of readily reusable software components,
and with their skills in managing hierarchical systems of complex software components with clear interfaces,
RSEs can now build reliable, maintainable systems of a scale and capability not previously possible.
\end{whatis}
